---
title: "Supplementary Data for 'Review and test of reproducibility of sub-decadal resolution palaeoenvironmental reconstructions from microfossil assemblages'"
author: "Richard J. Telford"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{float}
   - \floatplacement{figure}{H}
output: 
  bookdown::pdf_document2:
    keep_tex: true
    number_sections: false
    toc: false
bibliography: extra/chironomid2.bib
csl: extra/journal-of-paleolimnology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

options(digits = 2)
theme_set(theme_bw(base_size = 12))

Zabinskie <- "Żabińskie" #avoid upsetting spell checker with special characters
```

## Lake `r Zabinskie`

```{r zab-dist-analogue, fig.cap = "Squared chord distance from Lake Żabińskie fossil samples to nearest analogue in the Canadian-Polish calibration set. Analogue quality indicated by background colour estimated with the methods and thresholds used in the original paper. Calibration set samples with low count sums are omitted, but the results are similar if they are included."}
readd(zabinskie_dist_analogue_plot)
```

```{r zab-residlen, fig.cap = "Goodness of fit of the Lake Żabińskie fossil samples added passively to a CCA of the Canadian-Polish calibration set constrained by temperature."}
readd(zabinskie_residLen_plot)
```

## Abisko lakes

```{r abisko-all-lakes-plot, fig.cap = " Twentieth century reconstructed and instrumental July air temperatures for four lakes near Abisko. Data digitised from the original paper which reported that all the lakes had a similar correlation between the reconstructed and instrumental record."}
readd(abisko_all_lakes_plot)
```

## Lake Silvaplana
```{r silva-climate-plot, fig.cap="Instrumental July air temperatures from Selg-Maria [@Bergert2005] and the instrumental data as digitised from different papers. Digitised data are labelled after the calibration set used to make the reconstruction the instrumental data were compared against. All the digitised instrumental data should be identical, unless fossil samples have been merged/unmerged."}
readd(silva_climate_plot)
```

Long Swiss climate series were homogenised by Berget et al. [-@Bergert2005]. They were reprocessed in 2009 [@meteoswiss2010]. The Silvaplana papers probably all used the version of the Segl Maria series by Berget et al. [-@Bergert2005], which is shown in Fig. S\@ref(fig:silva-climate-plot). The differences between these data and the current version of the data are minimal (r^2^  = `r readd(silva_version_r2) %>% insist(. > 0.99) %>% format(digits = 3)`).

## Seebergsee

```{r seeberg-cit-plot, fig.cap = "Apparent and cross-validated performance of the calibration-in-time WAPLS-2 model from Seebergsee."}
readd(seeberg_cit_plot)
```

```{r seeberg-climate-plot, fig.cap = "Instrumental July air temperature anomalies from Château-d'Oex [@meteoswiss2010] and the instrumental data as digitised from the Seebergsee paper."}
readd(seeberg_climate_plot)
```

There may be, probably small, differences between the data shown in Fig. S\@ref(fig:seeberg-climate-plot) and the version of the Château-d'Oex temperature series homogenised by Berget et al. [@Berget2005].


## Nurmijärvi
```{r luoto-cit-plot, fig.cap = "Apparent and cross-validated performance of the calibration-in-time tolerance-weighted weighted averaging model from Nurmijärvi."}
readd(luoto_cit_plot)
```



## Speke Hall Lake
```{r speke-ana-dist-plot, fig.cap = "Squared chord distance from Speke Hall Lake fossil samples to nearest analogue in the Norwegian calibration set. Analogue quality indicated by background colour. Thresholds derived from quantiles of all calibration set distances (Good < 0.05, Fair < 0.1). "}
readd(speke_ana_dist_plot)
```

```{r speke-res-len-plot, fig.cap = "Goodness of fit of the Speke Hall Lake fossil samples added passively to a CCA of the Norwegian calibration set constrained by temperature. Thresholds derived from quantiles of all calibration set squared residual lengths (Good < 0.9, Fair < 0.95). An equivalent figure in the original paper used an inappropriate data tranformation prior to the analysis."}
readd(speke_resLen_plot)
```





