---
title: "Untitled"
author: "Richard J. Telford"
date: "May 21, 2016"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

It is conventional to [write small integers](http://www.apvschicago.com/2011/05/numbers-spell-out-or-use-numerals.html) as words rather than figures in text, especially if they are at the start of the sentence. This caused me some grief with [rmarkdown](http://rmarkdown.rstudio.com/), which I have started using for presentations, papers and blog-posts: my code gave me figures, I wanted words. Then I found the `english` package.

```{r}
library("english")
as.english(42)
as.english(-1)
```

We need to be careful if we use this inline with 
```{r, eval = FALSE}
The Ultimate Answer to Life, The Universe and Everything is...`r as.english(42)`
```
as this will be converted back to a figure.

> The Ultimate Answer to Life, The Universe and Everything is...`r as.english(42)`. 

Wrapping the command in `as.character` will fix this

```{r, eval = FALSE}
`r as.character(as.english(2))` to the power of `as.character(as.english(276709))` to `as.character(as.english(1))` against
```
> `r as.character(as.english(2))` to the power of `r as.character(as.english(276709))` to `r as.character(as.english(1))` against

It would be a cunning plan to make a function to run `as.character(as.english(x))` if it is needed repeatedly.

Of course, sometimes we will want to start a sentence with a number and need the first letter capitalising. This can be done with regular expressions (otherwise known as magic).

```{r, eval = FALSE}
`r gsub("(^)(.)", "\\1\\U\\2", as.english(42), perl = TRUE)`
```

> `r gsub("(^)(.)", "\\1\\U\\2", as.english(42), perl = TRUE)` 

As was hinted above by the conversion back to figures, objects made by `as.english` retain their numeric values, and so can be used in calculations 

```{r}
char.english <- function(x)as.character(as.english(x))
num <- lapply(0:9, as.english)
names(num) <- sapply(0:9, char.english)# set names
attach(num) #make available

six * nine
```